---
title: "Bayes Tutorial"
subtitle: "PSY 504: Advanced Statistics Final Project"
author: "Yeaju Diana Kim"
date: "2026-05-06"
date-modified: today
format:
  html:
    code-fold: true
    code-overflow: wrap
    toc: true
    toc-depth: 2
    toc-location: left
execute:
  message: false
  warning: false
engine: knitr
---

## Instructions

Here we are going to \~.

## Setup

```{r}
#| code-fold: false
library(tidyverse)
library(ggside)
library(brms)
library(broom)
library(broom.mixed)
library(ggdist)
library(tidybayes)
```

# Part 1: EDA, OLS, BRMS

## Loading data

First we load our data.

```{r}
# load data
data(penguins, package = "palmerpenguins")

# Any type of looking at data is a part of EDA
glimpse(penguins)
head(penguins)
```

Preprocess data.

```{r}
chinstrap <- penguins %>% 
  filter(species == "Chinstrap")

glimpse(chinstrap)
```

Now the data have \~ rows and \~.

## More Exploratory data analysis (EDA)

Our focal variables will be `body_mass_g` and `bill_length_mm`. Here they are in a scatter plot.

```{r}
chinstrap %>% 
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point() +
  stat_smooth(method = "lm", formula = 'y ~ x', se = FALSE)
```

We can augment the plot with some nice functions from the **ggside** package.

```{r}
chinstrap %>% 
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point() +
  stat_smooth(method = "lm", formula = 'y ~ x', se = FALSE) +
  # from ggside
  geom_xsidehistogram(bins = 30) +
  geom_ysidehistogram(bins = 30) +
  scale_xsidey_continuous(breaks = NULL) +
  scale_ysidex_continuous(breaks = NULL) +
  theme(ggside.panel.scale = 0.25)
```

It's a good idea to get a sense of the sample statistics. Here are the means and SD's for the two variables.

```{r}
chinstrap %>% 
  summarise(body_mass_g_mean = mean(body_mass_g),
            body_mass_g_sd = sd(body_mass_g),
            bill_length_mm_mean = mean(bill_length_mm),
            bill_length_mm_sd = sd(bill_length_mm)) 
```

And you know that more efficient way to compute sample statistics for multiple variables is to first convert the data into the long format with `pivot_longer()`. Then you use a `group_by()` line before the main event in `summarise()`.

```{r}
chinstrap %>% 
  pivot_longer(cols = c(body_mass_g, bill_length_mm)) %>% 
  group_by(name) %>% 
  summarise(mean = mean(value),
            sd = sd(value),
            # count the missing data (if any)
            n_missing = sum(is.na(value))) 
```

## OLS

We'll fit the model using brm(). We can also fit the model with the base **R** `lm()` function, which uses the OLS algorithm.

```{r}
# fit
fit1.ols <- lm(
  data = chinstrap,
  bill_length_mm ~ 1 + body_mass_g
)

# summarize the results
summary(fit1.ols)

coef(fit1.ols)
predict(fit1.ols)
predict(fit1.ols,
        interval = "confidence") %>% 
  # just the top 6
  head()

predict(fit1.ols,
        se.fit = TRUE) %>% 
  data.frame()

nd <- tibble(body_mass_g = seq(from = min(chinstrap$body_mass_g),
                               to = max(chinstrap$body_mass_g),
                               length.out = 50))

glimpse(nd)

predict(fit1.ols,
        interval = "confidence",
        newdata = nd) %>% 
  # just the top 6
  head()
```

Now we wrangle those predictions a bit and pump the results right into `ggplot()`.

```{r}
predict(fit1.ols,
        interval = "confidence",
        newdata = nd) %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  
  ggplot(aes(x = body_mass_g)) +
  # 95% confidence interval ribbon
  geom_ribbon(aes(ymin = lwr, ymax = upr),
              alpha = 1/3) +
  # point estimate line
  geom_line(aes(y = fit)) +
  geom_point(data = chinstrap,
             aes(y = bill_length_mm))

residuals(fit1.ols)

# put them in a tibble
tibble(r = residuals(fit1.ols)) %>% 
  # plot!
  ggplot(aes(x = r)) +
  geom_histogram(binwidth = 1)
```

## Bayes with default settings

We'll be fitting our Bayesian models with the **brms** package. The primary function is `brm()`.

`brm()` can work a lot like the OLS-based `lm()` function. For example, here's how to fit a Bayesian version of our OLS model `fit1.ols`.

```{r fit1.b, results = "hide"}
fit1.b <- brm(
  data = chinstrap,
  bill_length_mm ~ 1 + body_mass_g
)
```

Summarize the model.

```{r}
summary(fit1.b)
```

We can get a quick and dirty plot of the fitted line with the `conditional_effects()` function.

```{r}
conditional_effects(fit1.b) %>% 
  plot(points = TRUE)
```

## Coefficients and coefficient plots

We can compute a focused summary of the Bayesian model with the `fixef()` function.

```{r}
fixef(fit1.b)
```

If you would like to use the `tidy()` function with your **brms** models, it will have to be the version of `tidy()` from the **broom.mixed** package.

```{r}
tidy(fit1.b)
```

Here's how to wrangle and combine these two results into a single data frame. Then we'll make a coefficient plot.

```{r, warning = F}
bind_rows(
  tidy(fit1.ols, conf.int = TRUE) %>% select(term, estimate, contains("conf")),
  tidy(fit1.b) %>% select(term, estimate, contains("conf")) %>% filter(term != "sd__Observation")
) %>% 
  mutate(method = rep(c("lm()", "brm()"), each = 2)) %>% 
  
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = method)) +
  geom_pointrange() +
  scale_x_continuous("parameter space", expand = expansion(mult = 0.2)) +
  scale_y_discrete(expand = expansion(mult = 5)) +
  facet_wrap(~ term, scales = "free_x")
```

At a superficial level for simple conventional regression type models, the results from a Bayesian `brm()` model will be very similar to those from an OLS `lm()` model. This will not always be case, and even in this example there are many differences once we look below the surface.

# Part 2: Penguins and their posteriors

## Exploring model results

We can extract the posterior draws from our Bayesian models with the `as_draws_df()` function.

Note the meta data. We can get a sense of the full posterior distributions of the $\beta$ parameters with plots.

```{r}
as_draws_df(fit1.b)

# wrangle
as_draws_df(fit1.b) %>% 
  pivot_longer(starts_with("b_")) %>% 
  
  # plot!
  ggplot(aes(x = value)) + 
  # geom_density(fill = "grey20") +
  geom_histogram(bins = 40) +
  facet_wrap(~ name, scales = "free")
```

We might summarize those posterior distributions with basic descriptive statistics, like their means, SD's, and inner 95-percentile range.

```{r}
as_draws_df(fit1.b) %>% 
  pivot_longer(starts_with("b_")) %>% 
  group_by(name) %>% 
  summarise(mean = mean(value),
            sd = sd(value),
            ll = quantile(value, probs = 0.025),
            ul = quantile(value, probs = 0.975))
```

Notice how these values match up exactly with those from `fixef()`.

```{r}
fixef(fit1.b)
plot(fit1.b)
pairs(fit1.b)

# we can adjust some of the settings with the off_diag_args argument
pairs(fit1.b, off_diag_args = list(size = 1/4, alpha = 1/4))
```

Notice how the two $\beta$ parameters seem to have a strong negative correlation. We can quantify that correlation with the `vcov()` function.

```{r}
vcov(fit1.b)                      # variance/covariance metric
vcov(fit1.b, correlation = TRUE)  # correlation metric
```

## Draws

Let's save the `as_draws_df()` output for our model as an object called `draws`.

```{r}
draws <- as_draws_df(fit1.b)
glimpse(draws)
```

For each parameter in the model, we have 4,000 draws from the posterior. Let's break the 4000 draws down with our `draws` object.

```{r, warning = F}
# adjust the parameter names 
draws <- draws %>% 
  mutate(beta0 = b_Intercept,
         beta1 = b_body_mass_g)

# Note: go through this one line at a time
draws %>% 
  select(.draw, beta0, beta1) %>% 
  mutate(body_mass_g = mean(chinstrap$body_mass_g)) %>% 
  mutate(y_hat = beta0 + beta1 * body_mass_g) %>% 
  
  ggplot(aes(x = y_hat)) +
  geom_histogram(bins = 40) +
  labs(title = "Bayesians have posterior distributions",
       x = expression(hat(italic(y))*'|'*italic(x)==3733.1)) +
  coord_cartesian(xlim = c(47, 51))
```

Another handy way to present a Bayesian posterior is as a density with a point-interval summary below.

```{r, warning = F}
draws %>%
  mutate(body_mass_g = mean(chinstrap$body_mass_g)) %>% 
  mutate(y_hat = beta0 + beta1 * body_mass_g) %>% 
  
  ggplot(aes(x = y_hat)) +
  stat_halfeye(point_interval = mean_qi, .width = .95) +
  # scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Bayesians have posterior distributions",
       x = expression(hat(italic(y))*'|'*italic(x)==3733.1)) +
  coord_cartesian(xlim = c(47, 51))
```

The dot at the base of the plot is the posterior mean, and the horizontal line marks the 95% percentile-based interval. If you'd like to mark the median instead, set `point_interval = median_qi`. If you're like a different kind of horizontal interval, adjust the `.width` argument.

```{r, warning = F}
draws %>% 
  mutate(body_mass_g = mean(chinstrap$body_mass_g)) %>% 
  mutate(y_hat = beta0 + beta1 * body_mass_g) %>% 
  
  ggplot(aes(x = y_hat)) +
  # note the changes to this line
  stat_halfeye(point_interval = median_qi, .width = c(.5, .99)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Bayesians have posterior distributions",
       subtitle = "The dot marks the median.\nThe thicker line marks the 50% interval, and\nthe thinner line marks the 99% interval.",
       x = expression(hat(italic(y))*'|'*italic(x)==3733.1)) +
  coord_cartesian(xlim = c(47, 51))
```

## About those means, SD's, and intervals.

You can describe a Bayesian posterior in a lot of different ways. Earlier we said the posterior mean is the Bayesian point estimate. This isn't strictly true. Means are very popular, but you can summarize a posterior by its mean, median, or mode.

Let's see what this looks like in practice. First, we compute and save our statistics for each of our model parameters.

```{r, warning = F}
points <- draws %>% 
  rename(`beta[0]` = beta0,
         `beta[1]` = beta1) %>% 
  pivot_longer(cols = c(`beta[0]`, `beta[1]`, sigma), 
               names_to = "parameter") %>% 
  group_by(parameter) %>% 
  summarise(mean = mean(value),
            median = median(value),
            mode = Mode(value)) %>% 
  pivot_longer(starts_with("m"), names_to = "statistic")

# what?
points
```

Now plot.

```{r, warning = F}
draws %>% 
  rename(`beta[0]` = beta0,
         `beta[1]` = beta1) %>% 
  pivot_longer(cols = c(`beta[0]`, `beta[1]`, sigma), 
               names_to = "parameter") %>% 
  
  ggplot(aes(x = value)) +
  geom_density() +
  geom_vline(data = points,
             aes(xintercept = value, color = statistic),
             size = 3/4) +
  scale_color_viridis_d(option = "A", end = .8) +
  scale_y_continuous(NULL, breaks = NULL) +
  xlab("parameter space") +
  facet_wrap(~ parameter, labeller = label_parsed, scales = "free", ncol = 1) +
  theme(strip.text = element_text(size = 14))
```

-   The mean is the **brms** default summary, and McElreath (2015, 2020) defaulted to the mean in his texts.
-   The median is also available for many **brms** functions, and it's what Gelman et al (2020) recommend.
-   The mode can be attractive for very skewed distributions, and it's what Kruschke (2015) used in his text.

With many **brms** functions, you can request the median by setting `robust = TRUE`. For example:

```{r}
fixef(fit1.b)                 # means
fixef(fit1.b, robust = TRUE)  # medians
```

#### Bayesian intervals.

Bayesians describe the widths of their posteriors with intervals. I've seen these variously described as confidence intervals, credible intervals, probability intervals, and even uncertainty intervals. My recommendation is just pick a term, and clearly tell your audience what you mean (e.g., at the end of a Method section in a journal article).

Also, Bayesian intervals aren't always percentile based. An alternative is the highest posterior density interval (HPDI), which has mathematical properties some find desirable.

**brms** only supports percentile-based intervals, but it does allow for a variety of different ranges via the `prob` argument. For example, here's how to request 80% intervals in `summary()`.

```{r}
summary(fit1.b, prob = .80)
```

Regarding interval widths:

-   95% Intervals are widely used.
-   McElreat likes 89% intervals, and uses them as a default in his **rethinking** package.
-   Some of the **bayesplot**, **ggdist**, and **tidybayes** functions return 80% intervals.
-   Some of the **ggdist**, and **tidybayes** functions return 66% or 50% intervals.
-   I've heard Gelman report his fondness for 50% intervals on his blog (https://statmodeling.stat.columbia.edu/2016/11/05/why-i-prefer-50-to-95-intervals/).

Regarding interval types:

-   Percentile-based intervals are widely used in the Stan ecosystem, and are supported in texts like Gelman et al.
-   Kruschke has consistently advocates for HPDI's in his articles, and in his text.

### Posterior summaries with **tidybayes**.

Matthew Kay's **tidybayes** package (https://mjskay.github.io/tidybayes/) offers an array of convenience functions for summarizing posterior distributions with points and intervals. See the *Point summaries and intervals* section of Kay's *Extracting and visualizing tidy draws from brms models* vignette (https://mjskay.github.io/tidybayes/articles/tidy-brms.html#point-summaries-and-intervals) for a detailed breakdown. In short, the family of functions use the naming scheme `[median|mean|mode]_[qi|hdi]`. Here are a few examples.

```{r}
draws %>% mean_qi(beta0)                        # mean and 95% percentile interval
draws %>% median_qi(beta0, .width = .80)        # median and 80% percentile interval
draws %>% mode_hdi(beta0, .width = c(.5, .95))  # mode, with 95 and 50% HPDI's
```

### Spaghetti plots.

Remember how we said the `draw` was something like 4,000 separate equations for our Bayesian model? Let's see that again.

```{r}
draws %>% 
  select(.draw, beta0, beta1) %>% 
  mutate(body_mass_g = mean(chinstrap$body_mass_g)) %>% 
  # here's the equation
  mutate(y_hat = beta0 + beta1 * body_mass_g) %>% 
  # subset the top 6
  head()
```

One way we might emphasize the 4,000 equations is with a spaghetti plot. When we display the fitted line for `bill_length_mm` over the range of `body_mass_g` values, we can display a single line for each posterior draw. Here's what that can look like.

```{r, warning = F}
range(chinstrap$body_mass_g)

# Note: go through this one line at a time
draws %>% 
  select(.draw, beta0, beta1) %>% 
  expand_grid(body_mass_g = range(chinstrap$body_mass_g)) %>% 
  mutate(y_hat = beta0 + beta1 * body_mass_g) %>% 
  
  # plot!
  ggplot(aes(x = body_mass_g, y = y_hat, group = .draw)) +
  geom_line(linewidth = 1/10, alpha = 1/10)
```

While we're at it, let's take 20 draws and do a little color coding.

```{r, warning = F}
set.seed(20)

draws %>% 
  # take a random sample of 20 rows
  slice_sample(n = 20) %>% 
  select(.draw, beta0, beta1) %>% 
  expand_grid(body_mass_g = range(chinstrap$body_mass_g)) %>% 
  mutate(y_hat = beta0 + beta1 * body_mass_g) %>% 
  
  ggplot(aes(x = body_mass_g, y = y_hat, group = .draw, color = beta0)) +
  geom_line() +
  scale_color_viridis_c(expression(beta[0]~(the~intercept)), end = .9)
```

## Session information

```{r}
sessionInfo()
```
